---
title: "Stat 1 Final Review"
author: "Zhe Chen"
format: 
  html: default
  pdf: default
embed-resources: true
editor: visual
---

# Writing Part

## Chapter 1 Introduction

**Basic Equation**

$$
\text{DATA} = \text{MODEL} + \text{ERROR}
$$

One-parameter model:

$$
\begin{aligned}
Y_i &= \beta_0 + \text{ERROR} \\
    &= \beta_0 + \varepsilon_i \\
    &= \hat{Y_i} + e_i
\end{aligned}
$$

In general:

$$
\begin{aligned}
Y_i &= \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \cdots + \beta_j X_{ij} + \beta_{p-1} X_{i,p-1} + \varepsilon_i \\
    &= b_0 + b_1 X_{i1} + b_2 X_{i2} + \cdots + b_j X_{ij} + b_{p-1} X_{i,p-1} + e_i
\end{aligned}
$$

**Residual**: the part left over after using the model to predict the data

$$
\text{ERROR} = \text{DATA} - \text{MODEL}
$$

## Chapter 2: Simple Models - Errors

**Sum of Squared Errors** (**SSE**)

$$
\text{ERROR} = \sum_{i=1}^n e_i^2 = \sum_{i=1}^n (Y_i - \hat{Y_i})^2 
$$

In compact model:

$$
\text{ERROR} = \sum_{i=1}^n (Y_i - b_0)^2 = n(B_0-\bar{Y})^2
$$

**Mean** ($\bar{Y}$) minimizes the sum of square errors (SSE)

**Mean Squared Error** (**MSE**): the remaining error per potential parameter

$$
\text{MSE} = \frac{\text{SSE}} {n-p} = \frac{\sum_{i=1}^n {(Y_i - \hat{Y_i})^2}} {n-p} 
$$

For the simple model, MSE is **variance**

$$
\text{Variance} = s^2 = \frac{\text{SSE}} {n-1} = \frac{\sum_{i=1}^n {(Y_i - \hat{Y_i})^2}} {n-1} 
$$

**Standard Deviation** (**SD**) is the square root of MSE (**RMSE**). For the simple model:

$$
\text{Standard Deviation} = s = \sqrt{MSE} = \sqrt{\frac{\sum_{i=1}^n {(Y_i - \hat{Y_i})^2}} {n-1} }
$$

**Coefficient of Variation**(**CV**): the size of the standard deviation to be proportional to the size of the mean

$$
\text{Coefficient of Variation} = CV = \frac{s}{\bar{Y}}
$$

## Chapter 3: Simple Models - Error and Sampling Distributions

**Normal Distribution**

$$
X \sim N(\mu, \sigma^2)
$$

where $\mu$ is the expected value (mean) of the distribution, $\sigma^2$ is the variance of distribution. In standardized normal distribution, $\mu = 0, \sigma^2 = 1$.

In standardized normal distribution:

-   68% of observations fall between -1 and +1

-   95% of observations fall between -2 and +2

-   99.7% of observations fall between -3 and +3

**z score** (**standard normal deviates**):

$$
z = \frac{Y - \mu} {\sigma}
$$

**Assumption of Errors**

-   **Independence**: each error is independent of any other error
-   **Identically Distributed**: all errors are from the same distribution (same $\sigma^2$)
-   **Unbiased**: the expect value (mean) of the distribution from which the errors are sampled is zero

MSE ($s^2$) is an unbiased estimator of population error variance $\sigma^2$.

## Chapter 4: Simple Models - Statistical Inferences

**Compact Model**: Model C (***null hypothesis***)

**Augmented Model**: Mode A (***alternative hypothesis**)*

**Proportional Reduction in Error** (**PRE**)

$$ \begin{aligned} \text{PRE} &= \frac{\text{ERROR(C)} - \text{ERROR(A)}}{\text{ERROR(C)}} \\            &= 1 - \frac{\text{ERROR(A)}}{\text{ERROR(C)}} \end{aligned} $$

**Sum of Squares Reduced** (**SSR**):

$$
\text{SSR} = \text{SSE(C)} - \text{SSE(A)}
$$

$$
\text{PRE} = \frac{\text{SSR}}{\text{SSE(C)}}
$$

PRE overestimates and is a biased estimator of $\eta^2$.

**Statistic F**

$$
F = \frac{PRE/(PA-PC)}{(1-PRE)/(n-PA)}
$$

-   PA - PC: numerator degrees of freedom

-   n - PA: denominator degrees of freedom

-   numerator: average proportional reduction in error per parameter

-   denominator: average proportional reduction in error per parameter that could be added to the model

Critical values of $F (PA - PC = 1, \alpha = 0.05)$ :

-   n - PA = 10, F = 4.97

-   n - PA = 30, F = 4.17

-   n - PA = ∞, F = 3.84

Alternative formula of F:

$$
F = \frac{SSR/(PA-PC)}{(SSE(A))/(n-PA)} = \frac{MSR}{MSE}
$$

-   **MSR**: **mean squares reduced**

-   **MSE**: **mean square error**

**Statistic t**

$$ \sqrt{F_{1, n-1}} = t_{n-1} $$

## Chapter 5: Statistical Power

**Type I error** ($\alpha$): when Model C is correct, we incorrectly reject it ($\alpha$ = *significance level*)

**Type II error** *(*$\beta$) : when Model C is incorrect, we do not reject it

$1 - \beta$ : **statistical power:** *"*the likelihood that a statistical test leads the researcher to reject Model C *when Nature has determined that Model C is incorrect*."

**unbiased estimate** $\eta^2$:

$$
\eta^2 = 1 - (1-PRE)[\frac{n-PC}{n-PA}]
$$

**How to improve power:**

-   Increasing the size of the effect

-   Reducing error

-   Increasing $\alpha$

-   Increasing n

**Confidence Interval**

Definition (by S. Vaisey):

> If you were to repeat the entire data-generating-and-interval-building process indefinitely under identical conditions (same model, same population mean), then 99% of the intervals produced by this method would contain the true mean and 1% would not.

In the simple model:

$$
b_0 \pm \sqrt{\frac{F_{crit;1;n-1}MSE}{n}} = b_0 \pm t_{\alpha/2;n-1}SE
$$

**standard error** (**SE**):

$$
SE = \sqrt{\frac{MSE}{n}} = \frac{s}{\sqrt{n}}
$$

**Cohen's d**

$$
d = \frac{\beta_0 - B_0}{s} = 2 \sqrt{\frac{\eta^2}{1-\eta^2}}
$$

$$
\eta^2 = \frac{d^2}{d^2 + 4}
$$

## Section A: Recap

**Maximum Likelihood**: minimizing the sum of squared residuals is equivalent to maximizing the likelihood of the observations

## Chapter 6: Simple Regression

**Equation:**

$$
Y_i = \beta_0 + \beta_1X_i + \varepsilon_i
$$

**Calculate b1**

$$
b_1 = \frac{\sum(X_i - \bar{X})(Y_i - \bar{Y_i})}{\sum(X_i - \bar{X})^2} = \frac{s_{XY}}{s^2_X}
$$

**Calculate b0**

$$
b_0 = \bar{Y} - b_1\bar{X}
$$

**Pearson's r**

$$
r = \sqrt{PRE} \quad \text{or} \quad r^2 = PRE
$$

$$
r = \frac{\sum(X_i - \bar{X})(Y_i - \bar{Y_i})}{\sum(X_i - \bar{X})\sum(Y_i - \bar{Y})} = \frac{s_{XY}}{s_Xs_Y} = b_1\frac{s_X}{s_Y}
$$

**t**

$$
t = \frac{\beta_1}{SE}
$$

**Alternative Specifications**

-   **centering**

-   **normalization**

-   **x-standardization**

-   **full standardization**: fully standardized simple regression coefficient is exactly the same as **Pearson’s r** or the **correlation coefficient**

**Interpret Regression Results**

|             | Original | Centered | Norm.  | X-std. | XY-std. |
|:------------|:--------:|:--------:|:------:|:------:|:-------:|
| (Intercept) |  65.740  |  70.879  | 69.397 | 70.879 | -0.000  |
| hs_grad     |  0.097   |          |        |        |         |
| c_hs_grad   |          |  0.097   |        |        |         |
| hs_grad01   |          |          | 2.855  |        |         |
| z_hs_grad   |          |          |        | 0.782  |  0.582  |
| Num.Obs.    |    50    |    50    |   50   |   50   |   50    |
| F           |  24.615  |  24.615  | 24.615 | 24.615 | 24.615  |
| R2          |  0.339   |  0.339   | 0.339  | 0.339  |  0.339  |

> According to the model, **adjusting for** the murder rate, **each additional** percentage point of high school graduation **predicts** .044 more years of life expectancy.

> According to the model, **adjusting for** the high school graduation rate, **each unit increase** in the murder rate per 100,000 **predicts** .237 fewer years of life expectancy.

## Likelihood Comparison and Information Criteria

**Null Model:**

$$
Y_i \sim \mathcal{N}(\mu_i, \sigma) 
$$ $$
\mu_i = \beta_0
$$

-   A **distributional family** for the outcome
-   A **linear predictor** that links conditional expectations to the key parameter in that family (here, $\mu$, or the **conditional mean**)

**Likelihood Ratio Test**

$$
G^2 = 2[\mathcal{l}(\hat{\theta_1)}-\mathcal{l}(\hat{\theta_0)}] \overset{\text{approx}}{\sim} \chi^2_{df}
$$

**Deviance**

$D$ equals to -2 times of log-likelihood, it means "**how far away are the predictions of this model from**

**the data?**"

$$
D_M = -2l_M
$$

**Proportional Reduction in Deviance**

$$
PRD = \frac{D_C - D_A}{D_C}
$$

**AIC (Akaike's Information Criterion)**

$$
AIC = D + 2k
$$

-   $D$: deviance

-   $k$: number of estimated parameters (e.g., 3 for a simple regression)

-   motivation: out-of-sample prediction (increasingly complex models do an increasingly good job of fitting the training (in-sample) data but do an increasingly bad job of fitting the test(out-of-sample) data. In fact, the deviance difference is 2 per excess parameter.)

**BIC (Bayesian Information Criterion)**

$$
BIC = D + k*logn
$$

-   $logn$: natural log of the sample size

-   motivation: which of the models that we’re considering is most likely to be the “true model” that generated the data

## Chapter 7: Multiple Regression

$$ Y_i = \beta_0 + \beta_1X_{1i} + \beta_2X_{2i} + ... + \beta_{p-1}X_{p-1i} +  \varepsilon_i $$

$$ b_j = \frac{\sum(X_{xj} - \bar{X})(Y_i - \bar{Y_i})}{\sum(X_{ij} - \bar{X})^2} $$

$$ b_0 = \bar{Y} - b_1\bar{X_j} - b_2\bar{X_j} -...-b_{p-1}\bar{X}_{p-1}$$

## Chapter 8: Moderated and Nonlinear Regression Models

**Additive Model**

$$
Y_i = \beta_0 + \beta_1(X_{i1}) + \beta_2(X_{i2}) + \varepsilon_i
$$

or,

$$
Y_i = \mathcal{N}(\mu_i, \sigma) \\ 
\mu_i = \beta_0 + \beta_1(X_{i1}) + \beta_2(X_{i2}) 
$$

**Interaction Model**

$$
Y_i = \beta_0 + \beta_1(X_{i1}) + \beta_2(X_{i2}) + \\
\beta_3(X_{i1})(X_{i2}) + \varepsilon_i
$$

or,

$$
Y_i = \mathcal{N}(\mu_i, \sigma) \\ 
\mu_i = \beta_0 + \beta_1(X_{i1}) + \beta_2(X_{i2}) + \\ 
\beta_3(X_{i1})(X_{i2})
$$

-   the $X_{i1}$ null hypothesis is “the slope of $X_{i1}$ is 0 when $X_{i2}$ is 0”
-   the $X_{i2}$ null hypothesis is “the slope of $X_{i2}$ is 0 when $X_{i1}$ is 0”

**Polynomial Term**

-   $\beta_1X + \beta_2X^2$

-   the slope of $X$ depends on its own value

**Anova vs. Likelihood Ratio Test**

Likelihood ratio test gives a smaller p-value because the tail of the $\chi^2$ distribution is thinner than the tail of the $F$ distribution when the sample size is small. These get more similar as the sample size increases.

## Chapter 9-11: Contrast Coding, ANOVA, and ANCOVA

**Contrast Coding**

## Poisson and Chapter 15 Logistic Regression

**Poisson Distribution**

Poisson distribution has **only one parameter** $\lambda$, for both the **mean** and **variance**. Poisson distribution is for **count variables**, and a discrete probability distribution that assigns a finite probability to all positive integers using a probability mass function.

**Poisson Model**

$$
y_i = Poisson(\lambda_i) 
$$ $$
log(\lambda_i) = \beta_0 + \beta_1x_1
$$

or,

$$
\lambda_i = e^{\beta_0}e^{\beta_1x_i} = e^{\beta_0+\beta_1x_i}
$$

**Quantifying Difference**

1.  difference in probabilities

$$
p_1 - p_2
$$

2.  relative risk ratio (RRR)

$$
\frac{p_1}{p_2}
$$

3.  odds ratio (OR)

$$
\frac{p_1/1-p_1}{p2/1-p_2}
$$

odds = $\frac{p}{1-p}$

3.  log odds ratio

$$
\beta = log(\frac{p_1/1-p_1}{p2/1-p_2})
$$

the log odds ratio:

-   centered at **0** when there’s no association (instead of 1 for the OR)

-   Is **symmetric**: flipping which group is the reference just changes the sign

-   Ranges over all of $(-\infty, +\infty)$.

**Logit**

$$
logit(p) = log(\frac{p}{1-p}) = \beta
$$

$$
p = \frac{e^\beta}{1+e^\beta}
$$

# Coding Part

```{r}
#| message: false 

library(tidyverse)
library(ggplot2)
library(car)
library(here)
gss2024 <- readRDS(file = here::here("data", "gss2024.rds"))
```

## Basic `dplyr`

-   `filter()`

-   `select()`

-   `mutate()`

-   `summarize()`

-   `group_by()`

## Basic plotting

### Bar Chart

`geom_bar()`

```{r}
ggplot(data = cars,
       mapping = aes(x = speed)) +
  geom_bar()
```

### Histogram

`geom_histogram()`

```{r}
ggplot(data = cars,
       mapping = aes(x = speed)) +
  geom_histogram()
```

### Scatter Plot

`geom_point()`

```{r}
ggplot(data = cars,
       mapping = aes(x = speed,
                     y = dist)) +
  geom_point()
```

## SSE_A, SSE_C, and PRE

```{r}
d <- gss2024 |> 
  select(polviews, sex) |> 
  drop_na() |> 
  haven::zap_labels()
```

### SSE_A

```{r}
d <- d |> 
  mutate(polviews_dev = polviews - 4)

mod_a <- lm(polviews_dev ~ 1,
            data = d)
deviance(mod_a)
```

### SSE_C

```{r}
mod_c <- lm(polviews_dev ~ 0, 
            data = d)
deviance(mod_c) 
```

### PRE

```{r}
pre <- (deviance(mod_c) - deviance(mod_a)) / deviance(mod_a)
pre
```

## Statistic F and t

Get critical value:

-   `fcrit = qf(p, 1, df)`

-   `tcrit = qt((1 + p)/2, df)`

Get p value (right tail):

-   `p(f) = 1 - pf(fstat, 1, df)`

-   `p(t) = 2 * (1 - pt(sqrt(fstat), df))`

## Statistical Power

Command: `power.t.test()`

Example:

```{r}
# minimum interesting change (choice increase or decrease of 30 minutes)
delta <- .5

# estimate of variance of differences
sd_diff <- sd(d$polviews_dev)  # from GSS

# plug everything into the formula
power.t.test(delta = delta,
             sd = sd_diff,
             type = "one.sample", # I am comparing one sample's difference score
             sig.level = .01,
             power = .8)
```

For fully standardized variables: `pwr::pwr.r.test`

```{r}
pwr::pwr.r.test(r = .1, 
                sig.level = .01, 
                power = .9, 
                alternative = "two.sided")
```

## Likelihood Comparison

**Calculate likelihood of the model**

`logLik()`

```{r}
# compact model 
mc <- lm(polviews_dev ~ 0,
            data = d)
logLik(mc)

# augmented model 
ma <- lm(polviews_dev ~ 1,
            data = d)
logLik(ma)
```

**Likelihood ratio test**

`lmtext::lrtest()`

```{r}
lmtest::lrtest(mc, ma)
```

**AIC & BIC**

```{r}
AIC(mc, ma)
BIC(mc, ma)
```

## Simple & Multiple Regression

```{r}
#| message: false

state.x77_with_names <- as_tibble(state.x77,
                                  rownames = "state")
  
states <- bind_cols(state.x77_with_names, 
                    region = state.division) |> # both are in base R
  janitor::clean_names() |>                     # lower case and underscore
  mutate(south = as.integer(                    # make a south dummy
    str_detect(as.character(region), 
               "South")))
```

**Pairs Plot**

```{r}
states |> 
  select(life_exp, hs_grad, murder) |> 
  pairs()
```

**Correlation Matrix**

```{r}
states |> 
  select(life_exp, hs_grad, murder) |> 
  cor()
```

**Estimate Model**

```{r}
mod <- lm(life_exp ~ 1 + hs_grad + murder,
          data = states)
broom::tidy(mod)
broom::glance(mod)
```

**Transformation**

-   centering: `x - mean(x)`

-   normalization: `(x - min(x)) / (max(x) - min (x))`

-   standardization: `(x - mean(x)) / sd(x)`

## Interaction Models

```{r}
#| message: false 

state.x77_with_names <- as_tibble(state.x77,
                                  rownames = "state")
  
states <- bind_cols(state.x77_with_names, 
                    region = state.division) |> # both are in base R
  janitor::clean_names() |>                     # lower case and underscore
  mutate(south = as.integer(                    # make a south dummy
    str_detect(as.character(region), 
               "South"))) |> 
  mutate(inc1000 = income/1000)                 # income variable
```

**Additive Model**

```{r}
m1 <- lm(life_exp ~ 1 + hs_grad + inc1000,
         data = states)
broom::tidy(m1)
```

**Interaction Model**

```{r}
m2 <- lm(life_exp ~ 1 + hs_grad * inc1000, 
         data = states)
broom::tidy(m2)
```

**Visualization (continuous \* continuous)**

```{r}
# all values of hs_grad (x axis)
myhs <- seq(min(states$hs_grad),
            max(states$hs_grad),
            length = 100)

# representative values of income 
myincs <- quantile(states$inc1000,
                   c(0, 0.5, 1))


# all combinations 
xgrid <- expand_grid(inc1000 = myincs,
                     hs_grad = myhs)

# generate predictions
yhat <- predict(m2,
                newdata = xgrid,
                interval = "confidence",
                level = 0.95)

# bind together 
preds <- bind_cols(yhat, xgrid)

# convert inc1000 into factors 
preds <- preds |>
    mutate(income = case_match(inc1000,
                               min(states$inc1000) ~ "min",
                               max(states$inc1000) ~ "max",
                               .default = "median"))


# plotting 
ggplot(preds, aes(x = hs_grad, y = fit, fill = income, color = income)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2, color = NA) 
```

**Visualization (continuous \* categorical) with `ggpredict()`**

```{r}
# make sure categorical variables are factors 
states$south <- as.factor(states$south)

# interaction model
m3 <- lm(life_exp ~ 1 + hs_grad * south, 
         data = states)
broom::tidy(m3)


preds <- ggeffects::ggpredict(m3, terms = c("hs_grad", "south"))

plot(preds)
```

## T test

`t.test()`

```{r}
#| message: false

gss2024 <- readRDS("data/gss2024.rds")

d <- gss2024 |> 
  select(tvhours, degree, educ) |> 
  drop_na() |>
  haven::zap_labels() |> 
  mutate(college_fac = if_else(degree > 3, "BA+", "No BA"),
         college = if_else(degree > 3, 1, 0),
         degree_fac = factor(
           case_match(degree,
                      0 ~ "None",
                      1 ~ "HS",
                      2 ~ "AA",
                      3 ~ "BA",
                      4 ~ "GradProf"),
           levels = c("None", "HS", "AA", "BA", "GradProf")))
```

```{r}
t.test(tvhours ~ college,
       data = d,
       var.equal = TRUE) # assumes one sigma for both groups
```

## ANOVA

`anova()`

```{r}
anova(lm(tvhours ~ 1 + college,
         data = d))
```

## Contingency Table

`janitor::tabyl()`

```{r}
#| message: false

d <- readRDS("data/gss2024.rds") |> 
  haven::zap_labels() |> 
  select(tvhours, degree, sex) |> 
  drop_na() |> 
  mutate(sex = factor(sex, 
                      labels = c("M", "F")),        # sex as a factor
         degree_fac = factor(degree,
                             labels = c("none", "HS", "AA", "BA", "grad")))
```

```{r}
d |> 
  janitor::tabyl(sex, degree_fac) 
```

## Generalized Linear Models

```{r}
#| message: false 
d <- readRDS("data/gss2024.rds") |> 
  haven::zap_labels() |> 
  select(tvhours, degree, sex) |> 
  drop_na() |> 
  mutate(sex = factor(sex, 
                      labels = c("M", "F")),        # sex as a factor
         degree_fac = factor(degree,
                             labels = c("none", "HS", "AA", "BA", "grad"))) |>
    mutate(college_fac = if_else(degree > 3, "BA+", "No BA"))
```

### Linear (Gaussian)

```{r}
m1 <- glm(tvhours ~ 1 + sex,
          data = d,
          family = gaussian())
summary(m1)
```

### Poisson

```{r}
m2 <- glm(tvhours ~ 1 + sex,
          data = d,
          family = poisson())
summary(m2)
```

### Logistic

```{r}
m3 <- glm(degree_fac ~ 1 + sex,
          data = d,
          family = binomial(link = "logit"))
summary(m3)
```

#### Average Comparisons

Returns the **average change in predictions when a variable moves between values.**

```{r}
## risk difference 
marginaleffects::avg_comparisons(m3)

## risk ratio
marginaleffects::avg_comparisons(m3, comparison = "ratio")
```

#### Average Predictions

Returns the **average predicted outcome**. “What is the expected value of Y?”

```{r}
marginaleffects::avg_predictions(m3)
```

#### **Average Marginal Effects**

Returns the **average marginal effect (derivative)** of a variable. “How much does Y change for one unit change in X?”

```{r}
marginaleffects::avg_slopes(m3, variables = c("sex"))
```
