---
title: "Chapter 6 homework"
format:
  html:
    embed-resources: true
author: Zhe Chen
date: 2026-02-20
---

## Setup

```{r}
#| message: FALSE
library(tidyverse)
library(tinyplot)
library(here)
```

## Data

This homework uses a built-in dataset from base R (`state.x77`). Run the code below to create the working tibble `states`.

```{r}
state.x77_with_names <- as_tibble(state.x77,
                                  rownames = "state")

states <- bind_cols(state.x77_with_names,
                    region = state.division) |>
  janitor::clean_names() |>
  mutate(
    south = as.integer(
      str_detect(as.character(region), "South")
    )
  )
```

Throughout, use:

-   outcome: `life_exp`
-   predictor: `murder`

# Question 1: Descriptive statistics

## 1a. Descriptive statistics (psych)

Use `psych::describe()` (or something else if you want) to get descriptive statistics for `life_exp` and `murder`.

```{r}
states |>
  select(life_exp, murder) |>
  psych::describe()
```

## 1b. Descriptive statistics (tidyverse)

Compute (at minimum) the mean and standard deviation of `life_exp` and `murder` using `summarize()`.

```{r}
states |>
  select(life_exp, murder) |>
  summarize(
    mean_life_exp = mean(life_exp),
    sd_life_exp = sd(life_exp),
    mean_murder = mean(murder),
    sd_murder = sd(murder)
  )
```

# Question 2a: Regression and model comparison

You will evaluate whether `murder` predicts `life_exp`.

## 2a(i). Write the model comparison and null hypothesis

Write Model C and Model A in words and equations, and state the null hypothesis.

A typical setup for testing whether a slope is needed is:

-   Model C (constrained): slope fixed at 0
-   Model A (augmented): slope estimated from the data

Use the notation below (fill in the missing pieces in your write-up):

Model C:

$$
Y_i = \;\;\;\;\;\;\;\; + \varepsilon_i
$$

Model A:

$$
Y_i = \;\;\;\;\;\;\;\; + \;\;\;\;\;\;\;\; X_i + \varepsilon_i
$$

Null hypothesis (in symbols and words):

$$
H_0: \;\;\;\;\;\;\;\;
$$

------------------------------------------------------------------------

Model C: Murder rate is not associated with life expectancy at the state level.

$$
Life\_exp_i = \beta_0 + \varepsilon_i
$$

Model A: Murder rate is associated with life expectancy at the state level.

$$
Life\_exp_i = \beta_0 + \beta_1Murder_i + \varepsilon_i
$$

Null hypothesis: The slope of murder rate is 0.

$$
H_0: \beta_1 = 0
$$

## 2a(ii). Fit both models

```{r}
mc <- lm(life_exp ~ 1, data = states)
ma <- lm(life_exp ~ 1 + murder, data = states)
```

## 2a(iii). Compute SSE(C), SSE(A), PRE, and F

Compute:

-   $\mathrm{SSE}_C$ and $\mathrm{SSE}_A$ from the fitted models
-   $\mathrm{PRE} = (\mathrm{SSE}_C - \mathrm{SSE}_A)/\mathrm{SSE}_C$
-   the model-comparison $F$ statistic

Use `deviance()` for SSE (RSS) and `df.residual()` as needed.

$$
F = \frac{(\mathrm{SSE}_C - \mathrm{SSE}_A)/(P_A - P_C)}{\mathrm{SSE}_A/(n - P_A)}.
$$

```{r}
# SSE_C
sse_c <- deviance(mc)
sse_c

# SSE_A
sse_a <- deviance(ma)
sse_a

# PRE
pre <- (sse_c - sse_a) / sse_c
pre

# F
pc <- nrow(states) - df.residual(mc)
pc

pa <- nrow(states) - df.residual(ma)
pa

n <- nrow(states)
n

f <- ((sse_c - sse_a) / (pa - pc)) / (sse_a / (n - pa))
f
```

## 2a(iv). Write the prediction equation and interpret coefficients

```{r}
# model summary
summary(ma)

t <- -sqrt(f)
t

pf <- pf(f, 1, 48, lower.tail = FALSE)
pf

pt <- 2 * pt(t, 48)
pt
```

Write the prediction equation implied by Model A:

$$
\widehat{Life\_exp} = 72.97 - 0.28(Murder)
$$

Interpret:

-   the intercept:
    -   When the state's murder rate is 0%, its average life expectancy is 72.97 years.
-   the slope:
    -   For each percentage increase in the state's murder rate, the average life expectancy decreases by 0.28 years.

## 2a(v). Conclusion

State:

-   statistical conclusion (report $F$ or $t$, PRE, and $p$):
    -   By moving from Model C to Model A, the PRE is 0.61, with the $F$ is 74.99 ($t$ = -8.66), $p$ = 2.26e-11. In other words, given the null hypothesis is true, the observed $F$ is very surprising. Therefore, we should reject the null hypothesis.
-   substantive conclusion (in plain language)
    -   Murder rate is negatively associated with average life expectancy at the state level. For each percentage increase in the state's murder rate, its average life expectancy decreases by 0.28 years.

# Question 2b: Center the predictor

Create a centered version of murder:

## 2b(i). Create `murder_c` and refit the model

```{r}
states <- states |>
  mutate(murder_c = murder - mean(murder))

mc_c <- lm(life_exp ~ 1, data = states)
ma_c <- lm(life_exp ~ 1 + murder_c, data = states)
```

## 2b(ii). Report SSE(A), SSE(C), PRE, and F again

Compute the same fit statistics (SSE, PRE, F). Compare them to the results from Question 2a.

```{r}
# SSE_C
sse_c_c <- deviance(mc_c)
sse_c_c

# SSE_A
sse_a_c <- deviance(ma_c)
sse_a_c

# PRE
pre_c <- (sse_c_c - sse_a_c) / sse_c_c
pre_c

# F
pc_c <- nrow(states) - df.residual(mc_c)
pc_c

pa_c <- nrow(states) - df.residual(ma_c)
pa_c

n_c <- nrow(states)
n_c

f_c <- ((sse_c_c - sse_a_c) / (pa_c - pc_c)) / (sse_a_c / (n_c - pa_c))
f_c

# compare with 2a
all.equal(sse_c, sse_c_c)
all.equal(sse_a, sse_a_c)
all.equal(pre, pre_c)
all.equal(f, f_c)
```

The SSE_C, SSE_A, PRE, F are the same as those in 2a.

## 2b(iii). Interpret coefficients after centering

```{r}
# model summary
summary(ma_c)

t_c <- -sqrt(f_c)
t_c

pf_c <- pf(f_c, 1, 48, lower.tail = FALSE)
pf_c

pt_c <- 2 * pt(t_c, 48)
pt_c
```

Write the prediction equation for the centered model and interpret the intercept and slope.

In particular: what does the intercept represent *now*?

$$
\widehat{Life\_exp} = 70.88 - 0.28(Murder\_c)
$$

Interpret:

-   the intercept:
    -   When the state's murder rate is the mean value (7.38), its average life expectancy is 70.88 years.
-   the slope:
    -   For each percentage increase in the state's murder rate, the average life expectancy decreases by 0.28 years.

## 2b(iv). Plot and compare

Make two scatterplots with regression lines:

1.  `life_exp` vs. `murder`
2.  `life_exp` vs. `murder_c`

How are these plots different?

```{r}
par(mfrow = c(1, 2))

# life_exp vs. murder
tinyplot(life_exp ~ murder, data = states,
          xlim = c(0, max(states$murder)))
abline(lm(life_exp ~ murder, data = states), col = "red")

# life_exp vs. murder_c
tinyplot(life_exp ~ murder_c, data = states)
abline(lm(life_exp ~ murder_c, data = states), col = "red")
```

These two plot have the same slopes, as the the $b_1$ in 2a and 2b is the same. However, the intercepts are different. In `life_exp` vs. `murder`, the intercept indicates the prediction of average life expectancy when the murder rate is 0; while in `life_exp` vs. `murder_c`, the intercept indicates the prediction of average life expectancy when the murder rate is the mean value.

# Question 2c: Change the units of the predictor

Create a transformed predictor by standardizing `murder`.

## 2c(i). Create `murder_std` and refit the model

```{r}
states <- states |>
  mutate(murder_std = (murder - mean(murder)) / sd(murder))

mc_std <- lm(life_exp ~ 1, data = states)
ma_std <- lm(life_exp ~ 1 + murder_std, data = states)
```

## 2c(ii). Report SSE(A), SSE(C), PRE, and F again

Compute SSE, PRE, and F again and compare them to 2a and 2b.

```{r}
# SSE_C
sse_c_std <- deviance(mc_std)
sse_c_std

# SSE_A
sse_a_std <- deviance(ma_std)
sse_a_std

# PRE
pre_std <- (sse_c_std - sse_a_std) / sse_c_std
pre_std

# F
pc_std <- nrow(states) - df.residual(mc_std)
pc_std

pa_std <- nrow(states) - df.residual(ma_std)
pa_std

n_std <- nrow(states)
n_std

f_std <- ((sse_c_std - sse_a_std) / (pa_std - pc_std)) / (sse_a_std / (n_std - pa_std))
f_std

# compare with 2a, 2b
all.equal(sse_c, sse_c_c, sse_c_std)
all.equal(sse_a, sse_a_c, sse_a_std)
all.equal(pre, pre_c, pre_std)
all.equal(f, f_c, f_std)
```

The SSE_C, SSE_A, PRE, F are the same as those in 2a and 2b

## 2c(iii). Interpret coefficients after rescaling

Write the prediction equation and interpret the slope carefully in the new units.

```{r}
# model summary
summary(ma_std)

t_std <- -sqrt(f_std)
t_std

pf_std <- pf(f_std, 1, 48, lower.tail = FALSE)
pf_std

pt_std <- 2 * pt(t_std, 48)
pt_std
```

$$
\widehat{Life\_exp} = 70.88 - 1.05(Murder\_std)
$$

Interpret:

-   the intercept:
    -   When the state's murder rate is mean value (7.38) , its average life expectancy is 70.88 years.
-   the slope:
    -   For each standard deviation increase in the state's murder rate, the average life expectancy decreases by 1.04 years.

# Question 3: Predicted values and what transformations do (and do not) change

You now have three versions of the same basic relationship:

1.  using `murder`
2.  using `murder_c`
3.  using `murder_std`

## 3a. Predicted life expectancy at a specific murder rate

Using each of your three fitted Model A versions, compute the predicted life expectancy for a state with `murder` = 8.

Show your work. (Be explicit about how you convert 8 into the appropriate value for `murder_c` and `murder_std`.)

```{r}
# using murder
summary(ma)
life_exp <- 72.97356 - 0.28395 * 8
life_exp

# using murder_c
summary(ma_c)
life_exp_c <- 70.87860 - 0.28395 * (8 - mean(states$murder))
life_exp_c

# using murder_std
summary(ma_std)
life_exp_std <- 70.8786 - 1.048 * ((8 - mean(states$murder))/ sd(states$murder))
life_exp_std

# check 
all.equal(life_exp, life_exp_c, life_exp_std)
```

## 3b. Summary paragraph

Write a brief summary explaining what you learned about transforming a predictor:

-   How do these transformations affect the estimated intercept and slope?
-   Do they change predicted values (when transformed consistently)?
-   Do they change overall goodness of fit (SSE, PRE, $F$)?

Summary:

-   The transformation from using `murder` to `murder_c` does not change the slope, but changes the intercept; the transformation from using `murder_c` to `murder_std` does not change the intercept, but changes the slope; the transformation from using `murder` to `murder_std` changes both the intercept and the slope.
-   No, they do not change predicted values.
-   No, they do not change overall goodness of fit (SSE, PRE, $F$).

# Project application

Do the same thing with an outcome and predictor of your choice from your own dataset. Use three transformations:

-   original
-   normalized X
-   fully standardized

Interpret the intercepts and slopes for each of the three models. Compare the fit statistics for each of the three models.

```{r}
gss2024 <- readRDS(file = here::here("data", "gss2024.rds"))
# tvhours: hours per day watching tv
# age: age of respondent
gss <- gss2024 |> 
  select(tvhours, age) |>
  drop_na() |>
  haven::zap_labels()
psych::describe(gss)
```

Model C: Age is not associated with hours per day watching TV.

$$
\text{TVhours}_i = \beta_0 + \varepsilon_i
$$ Model A: Age is associated with hours per day watching TV.

$$
\text{TVhours}_i = \beta_0 + \beta_1\text{Age}_i + \varepsilon_i
$$

Null hypothesis: The slope of age is 0.

$$
\beta_1 = 0
$$

## Original

```{r}
# fit models
mc <- lm(tvhours ~ 1, data = gss)
ma <- lm(tvhours ~ 1 + age, data = gss)

# SSE_C
sse_c <- deviance(mc)
sse_c

# SSE_A
sse_a <- deviance(ma)
sse_a

# PRE
pre <- (sse_c - sse_a) / sse_c
pre

# F
pc <- nrow(gss) - df.residual(mc)
pc

pa <- nrow(gss) - df.residual(ma)
pa

n <- nrow(gss)
n

f <- ((sse_c - sse_a) / (pa - pc)) / (sse_a / (n - pa))
f

# summary Model A
summary(ma)
```

Prediction equation:

$$
\widehat{\text{TVhours}} = 1.46 + 0.04 (\text{Age})
$$

Interpret:

-   the intercept:
    -   When the individual's age years are 0, she watches TV for 1.46 hours per day.
-   the slope:
    -   For each year increase in the individual's age, her hours watching TV per day increase by 0.04 hours.

## Normalized X

```{r}
# normalize X 
gss <- gss |>
  mutate(age_n = (age - min(age)) / (max(age) - min(age)))
psych::describe(gss)

# fit models
mc_n <- lm(tvhours ~ 1, data = gss)
ma_n <- lm(tvhours ~ 1 + age_n, data = gss)

# SSE_C
sse_c_n <- deviance(mc_n)
sse_c_n

# SSE_A
sse_a_n <- deviance(ma_n)
sse_a_n

# PRE
pre_n <- (sse_c_n - sse_a_n) / sse_c_n
pre_n

# F
pc_n <- nrow(gss) - df.residual(mc_n)
pc_n

pa_n <- nrow(gss) - df.residual(ma_n)
pa_n

n <- nrow(gss)
n

f_n <- ((sse_c_n - sse_a_n) / (pa_n - pc_n)) / (sse_a_n / (n - pa_n))
f_n

# summary Model A
summary(ma_n)
```

Prediction equation:

$$
\widehat{\text{TVhours}} = 2.12 + 2.61 (\text{Age})
$$

Interpret:

-   the intercept:
    -   When the individual's age is the least value (18 years), she watches TV for 2.12 hours per day.
-   the slope:
    -   The individual with the oldest age (89 years) has 2.61 more hours watching TV per day, than the individual with the youngest age (18 years).

## Fully Standardized

```{r}
# standardize X and Y 
gss <- gss |>
  mutate(age_std = (age - mean(age)) / sd(age),
         tvhours_std = (tvhours - mean(tvhours)) / sd(tvhours))
psych::describe(gss)

# fit models
mc_std <- lm(tvhours_std ~ 1, data = gss)
ma_std <- lm(tvhours_std ~ 1 + age_std, data = gss)

# SSE_C
sse_c_std <- deviance(mc_std)
sse_c_std

# SSE_A
sse_a_std <- deviance(ma_std)
sse_a_std

# PRE
pre_std <- (sse_c_std - sse_a_std) / sse_c_std
pre_std

# F
pc_std <- nrow(gss) - df.residual(mc_std)
pc_std

pa_std <- nrow(gss) - df.residual(ma_std)
pa_std

n <- nrow(gss)
n

f_std <- ((sse_c_std - sse_a_std) / (pa_n - pc_n)) / (sse_a_std / (n - pa_n))
f_std

# summary Model A
summary(ma_std)
```

Prediction equation:

$$
\widehat{\text{TVhours}} = 0 + 0.02 (\text{Age})
$$

Interpret:

-   the intercept:
    -   When the individual's age is the mean value (50.35 years), she watches TV for the average hours (3.32 hours) per day.
-   the slope:
    -   For each standard deviation increase in the individual's age, her hours watching TV per day increase by 0.02 standard deviation in hours.

## Compare Fit Statistics

```{r}
all.equal(sse_c, sse_c_n, sse_c_std)
all.equal(sse_a, sse_a_n, sse_a_std)
all.equal(pre, pre_n, pre_std)
all.equal(f, f_n, f_std)
```

Their fit statistics (SSE, PRE, $F$) are the same.
